When do the effects of interaction happen? Interacting and disturbed galaxies in the COSMOS field

### 2024-01-16 paper-source/main.tex
Major update to error calculation. Updated names of stages to reflect what they are.

% Abstract of the paper
\begin{abstract}
   Abstract will go here.
\end{abstract}

@@

Often, the stage of an interaction is approximated by measuring different parameters with the projected separation between two systems. Early works showed that there was significant star formation enhancement (SFE) in galaxy pairs with small separations \citep{2003MNRAS.346.1189L ,2008AJ....135.1877E, 2022ApJ...940....4S}. A connection was also found between the projected separation in galaxy pairs and the AGN fraction where, once again, the AGN fraction appears to increase with decreasing projected separation \citep{2009MNRAS.399.2172R, 2011MNRAS.418.2043E, 2017MNRAS.465.2671G, 2021ApJ...909..124S}. However, using projected separation as a proxy for the stage in dynamical time of an interacting history overlooks which part of the dynamical history the interaction is at. For instance if a galaxy pair is found to have small projected separation, without visually confirming the morphology, it is difficult to ascertain if the galaxies are just approaching each other or just passed each other. The differences between these being paramount to understanding the effects of interaction on the galaxies involved.

@@

Figure \ref{fig:agn-frac-time} shows the changing AGN fraction with interaction stage. We find that the AGN fraction generally increases with interaction stage. There is a large increase from stage 1 to 2, followed by the fraction mostly remaining unchanged until stage 3. Then, in stage 4 we measure another large jump in the AGN fraction. However, before we discuss this result further, we must point out that the large error bars on our measurements and the low number counts of confirmed AGN and SFG make this difficult to interpret. We measure the errors on our sample from bootstrapping the underlying low-count population of AGN and SFGs and comparing to the change in fraction measured. Thus, while our initial values show a general increase, because of the large error bars, there could be no increase at all with stage to even a slight decline.

@@

We find that the AGN fraction generally increases with interaction stage, however, it again remains constant from stage 2 to stage 3. This indicates that the mechanism driving AGN ignition in interaction is primarily at the point of the closest approach between the two systems. This is further reinforced when the system begins to merge, and we see the AGN fraction increase significantly in our sample. When we investigate the AGN number counts with respect to the projected separation of systems, we find that it gradually declines over the projected separation. However, we find two peaks; first clearly in stage 2 of the interaction where the galaxies are overlapping and secondly in the early parts of stage 3. This could be evidence of a delayed AGN ignition depending on the underlying parameters of the interaction. Thus, \textbf{we find that the AGN fraction generally increases with stage, however, the mechanisms responsible for an increase in the fraction primarily occur when the two systems are at pericentre or are actually merging. We find evidence of a delay of ignition, as there is an increase in fraction at a early projected separations in stage 3}.

@@

\subsection{Interaction Stage and AGN}
\noindent The evolution of the AGN fraction in interacting galaxies is similar to that of the evolution of star formation. It has often been found with projected separation the AGN fraction increases. There are multiple observational works that show this \citep{2007MNRAS.375.1017A, 2013MNRAS.435.3627E, 2020ApJ...904..107S} as well as works on cosmological simulations which support these conclusions \citep{2023MNRAS.519.4966B}. In simulations, the increased likelihood of nuclear ignition comes from the sudden increase in gas density in the galactic core which naturally leads to increased black hole accretion and ignition. Observations of interacting galaxies are often interesting, as they contain examples of duel AGN and individual examples \citep[e.g.][]{2017MNRAS.470L..49E, 2021ApJ...923...36S} or investigate the increase in the AGN fraction in only the merger / post-merger stage \citep{2020A&A...637A..94G}.

We find an increasing AGN fraction with both stage and projected separation, before the galaxy merging and coalescence takes place. This is supported with other works which specifically look at AGN fraction with projected separation \citep{2011MNRAS.418.2043E, 2023ApJ...942..107S}. However, we specifically find that the AGN fraction increases rapidly from stage 1 to stage 2 and stage 3 to stage 4. Again, it holds quite constant between stages 2 and 3. This shows, again, shows two distinct end stages of interaction which can colour our interpretation of the link between AGN and interaction. It appears, from our results, that the AGN fraction is driven up during the points in the dynamical time when interacting galaxies are actually flying past each other and being morphologically disturbed. There is evidence that the onset of nuclear ignition from interaction may be delayed \citep{2011MNRAS.418.2043E}, or even flicker \citep{2015MNRAS.451.2517S} after ignition. Both of these possibilities are reflected distributions of AGN fraction with stage and projected separation.

Again, it is important to note the rather `wide net' that our stage classification takes. There is overlap between our stage 2 and stage 3 classifications at high stage 2 projected separation and low stage 3 projected separations. However, these stages naturally lead onto one another. Therefore, what see in the bi-modal distribution of AGN fraction with projected separation in Figure \ref{fig:sfg-agn-proj} are two peaks. The second of these peaks is at the cross over point of stage 2 to stage 3; the point at which (if the two galaxies have just flown by each other) a delayed AGN ignition coult take place. This is also reflected in Figure \ref{fig:agn-stage} where we see a slight increase in AGN fraction in stage 3. This is likely from the ignition of nuclear activity taking place over a large period of time with a delay being involved from the initial flyby. Other works do also support this interpretation.

@@

% SFR and projected separation
To further investigate this change in enhancement, we investigate our sub-sample of galaxy pairs and compare it to a mass and redshift matched control sample. We bin our projected separations and measure the ratio between the average interacting SFR and control SFR in each bin. We find, across the whole subsample, a general increase in star formation efficiency as we move to smaller projected separation. The highest found enhancement is below 10kpc. However, when broken down into their consituent stages, we find dramatically different behaviour in the star formation efficiency. This is best seen in stage 2 enhancement, which does not seem to change with projected separation, while there is a dramatic decrease in stage 3 enhancement from 2.5 at 10kpc down to 1.2 at 100kpc. This shows that just using projected separation as a proxy for stage will leave out crucial information to the underlying causes and mechanisms fueling star formation enhancement. To confirm that the effects observed here are not related to biases in the environment, we investigate its relation to our staged sample. We find that the environment is consistent between all stages, with no biases existing.

% Stage and AGN
Finally, we investigate the change in AGN activity across our whole sample. We find that, on average, the AGN fraction increases with stage. This is also true when related to the projected separation, where we find an almost bi-modal distribution. This is from a higher than average AGN fraction existing in the stage 2 and stage 3 samples, likely driven by stochastic inflows of gas during the interaction. We also find that the AGN fraction is highest in merging and merged galaxies.

% What does it mean?
While the results with projected separation are not unexpected, those with the different stages very much are. We have shown that one cannot simply use the projected separation of interacting galaxies as a proxy for the stage ongoing. We find very different behaviour in the star forming behavior of interacting galaxies based on stage they are at in the dynamical timescale of the interaction. We find the beginning of an enhancement in star formation occurs in stage 2, where the two galaxies are moving past each other at pericentre. Then, through to stage 3 the enhancement remains although with no further increase. There is then another dramatic starburst in stage 4, where the two galaxies actually begin to undergo the merging processes. However, the results are not so clear for AGN. While we find, definitively, that the AGN fraction increases with stage we are plagued by low counts and selection effects in these results. This is due to the limited classifications available of active galactic nuclei in our sample.

% Future Work
Our work shows the importance of considering morphological stage when considering interaction, and that there is a fine interplay between underlying proesses and the point of the dynamical timescale. We require larger samples of correctly staged galaxies to further understand and exploit what these relations are, and how to best investigate when and where star formation and its enhancement occurs. This is also particularly true of the relation between active galactic nuclei and the interaction stage, where we are unable to have the sample size to definitively draw conclusions from our results.

>>>>

% Abstract of the paper
\begin{abstract}
   Galaxy interaction and merging have fundamental impacts on galactic evolution. From increases in star formation, to morphology, to activation of nuclear activity, galaxy interaction is thought to play a role. However, at what point in the interaction do these effects occur? In this work, we break down a large sample of identified interacting galaxies and split them into four distinct stages based on their interaction morphology. Each stage captures a different part of the dynamical timescale of interaction; from the systems approaching each other to actually merging. We cross match this large sample with the COSMOS field to provide ancillary data and each galaxys' underlying parameters. We then investigate the relation between stage, mass, star formation, and nuclear activation. Using Kolmogorov-Smirnov tests, we find that the distribution of star formation rates rapidly changes with stage for an equivalent mass distribution. We find a rapidly shrinking red sequence with stage, and a large increase in the fraction of starburst galaxies at the point of coalescence. We also find that the fraction of active galactic nuclei is static through most of the interaction, until it dramatically increases at coalescence. We discuss these results, and compare them with the more prevalent stage diagnostic of projected separation between galaxy pairs. We find interesting behaviour that is morphology and stage dependent in our samples star formation rates which is not captured by only using the projected separation. We show that morphological stage and, more specifically, the point in time of the underlying dynamical timescale an interaction occupies is as important to consider as its projected separation. We conclude this work with discussions on how this could be applied to larger, more complete datasets of interacting galaxies with available ancillary data.
\end{abstract}

@@

Often, the stage of an interaction is approximated by measuring different parameters with the projected separation between two systems. Early works showed that there was significant star formation enhancement (SFE) in galaxy pairs with small separations \citep{2003MNRAS.346.1189L,2008AJ....135.1877E, 2022ApJ...940....4S}. A connection was also found between the projected separation in galaxy pairs and the AGN fraction where, once again, the AGN fraction appears to increase with decreasing projected separation \citep{2009MNRAS.399.2172R, 2011MNRAS.418.2043E, 2017MNRAS.465.2671G, 2021ApJ...909..124S}. However, using projected separation as a proxy for the stage in dynamical time of an interacting history overlooks which part of the dynamical history the interaction is at. For instance if a galaxy pair is found to have small projected separation, without visually confirming the morphology, it is difficult to ascertain if the galaxies are just approaching each other or just passed each other. The differences between these being paramount to understanding the effects of interaction on the galaxies involved.

@@

Figure \ref{fig:agn-frac-time} shows the changing AGN fraction with interaction stage. We find that the AGN fraction generally remains static with stage. There is a small decrease increase from stage 1 to 2, followed by the fraction mostly remaining unchanged until stage 3. Then, in stage 4 we measure a large jump in the AGN fraction. However, before we discuss this result further, we must point out that the large error bars on our measurements and the low number counts of confirmed AGN and SFG make this difficult to interpret. We measure the errors as the standard deviation about the mean in each stage. Thus, while our initial values show relatively little change, because of the large error bars, there could be increase with stage to even a further decline.

@@

We find the AGN fraction only increases at the point of coalescence in interaction: stage 4. This indicates that the mechanism driving AGN ignition in interaction is primarily at the point of the closest approach between the two systems. When we investigate the AGN number counts with respect to the projected separation of systems, we find that it gradually declines over the projected separation. However, we find two peaks; first clearly in stage 2 of the interaction where the galaxies are overlapping and secondly in the early parts of stage 3. This could be evidence of a delayed AGN ignition depending on the underlying parameters of the interaction. Thus, \textbf{we find that the AGN fraction is generally unchanged with stage, however, the mechanisms responsible for an increase in the fraction primarily occur when the two systems are actually merging. We find evidence of a delay of ignition, as there is an increase in fraction at a early projected separations in stage 3}.

@@

\subsection{Interaction Stage and AGN}
\noindent The evolution of the AGN fraction in interacting galaxies is similar to that of the evolution of star formation. It has often been found with projected separation the AGN fraction increases. There are multiple observational works that show this \citep{2007MNRAS.375.1017A, 2013MNRAS.435.3627E, 2020ApJ...904..107S} as well as works on cosmological simulations which support these conclusions \citep{2023MNRAS.519.4966B}. In simulations, the increased likelihood of nuclear ignition comes from the sudden increase in gas density in the galactic core which naturally leads to increased black hole accretion and ignition. To satisfy these conditions, actual coalescence of the two systems are required. Observations of interacting galaxies are often interesting, as they contain examples of duel AGN and individual examples \citep[e.g.][]{2017MNRAS.470L..49E, 2021ApJ...923...36S} or investigate the increase in the AGN fraction in only the merger / post-merger stage \citep{2020A&A...637A..94G}.

We find peaks in the AGN fraction with projected separation, before the galaxy merging and coalescence takes place. This is supported with other works which specifically look at AGN fraction with projected separation \citep{2011MNRAS.418.2043E, 2023ApJ...942..107S}. However, we specifically find that the AGN fraction increases rapidly in stage 4 and holds quite constant between stages 1 to 3. This shows, again, shows two distinct effects of interaction which can colour our interpretation of the link between AGN and interaction. It appears, from our results, that the AGN fraction is driven up during the points in the dynamical time when the inner parts of the galaxy are majorly disturbed, and not by the simple movement of gas and dust into the core during the two galaxies passing each other. There is evidence that the onset of nuclear ignition from interaction may be delayed \citep{2011MNRAS.418.2043E}, or even flicker \citep{2015MNRAS.451.2517S} after ignition. Both of these possibilities are reflected distributions of AGN fraction with projected separation.

Again, it is important to note the rather `wide net' that our stage classification takes. There is overlap between our stage 2 and stage 3 classifications at high stage 2 projected separation and low stage 3 projected separations. However, these stages naturally lead onto one another. Therefore, what see in the bi-modal distribution of AGN fraction with projected separation in Figure \ref{fig:sfg-agn-proj} are two peaks. The second of these peaks is at the cross over point of stage 2 to stage 3; the point at which (if the two galaxies have just flown by each other) a delayed AGN ignition could take place. This is also reflected in Figure \ref{fig:agn-stage} where we see a slight increase in AGN fraction in stage 3. This is likely from the ignition of nuclear activity taking place over a large period of time with a delay being involved from the initial flyby.

@@

% SFR and projected separation
To further investigate this change in enhancement, we investigate our sub-sample of galaxy pairs and compare it to a mass and redshift matched control sample. We bin our projected separations and measure the ratio between the average interacting SFR and control SFR in each bin. We find, across the whole subsample, a general increase in star formation efficiency as we move to smaller projected separation. The highest found enhancement is below 10kpc. However, when broken down into their constituent stages, we find dramatically different behaviour in the star formation efficiency. This is best seen in stage 2 enhancement, which does not seem to change with projected separation, while there is a dramatic decrease in stage 3 enhancement from 2.5 at 10kpc down to 1.2 at 100kpc. This shows that just using projected separation as a proxy for stage will leave out crucial information to the underlying causes and mechanisms fueling star formation enhancement. To confirm that the effects observed here are not related to biases in the environment, we investigate its relation to our staged sample. We find that the environment is consistent between all stages, with no biases existing.

% Stage and AGN
Finally, we investigate the change in AGN activity across our whole sample. We find that, on average, the AGN fraction remains constant with stage until the point of coalescence. However, when looking at project separation, we find an almost bi-modal distribution. This could be evidence for a delayed ignition in the AGN present in those systems undergoing an interaction. However, it is difficult to draw this conclusion definitively due to small number counts. We also find that the AGN fraction is highest in merging and merged galaxies.

% What does it mean?
While the results with projected separation are not unexpected, those with the different stages are. We have shown that one cannot simply use the projected separation of interacting galaxies as a proxy for the stage of the interaction. We find very different behaviour in the star forming behavior of interacting galaxies based on stage which are at in the dynamical timescale of the interaction. We find the beginning of an enhancement in star formation occurs in stage 2, where the two galaxies are moving past each other at pericentre. Then, through to stage 3, the enhancement remains although with no further increase. There is then another dramatic starburst in stage 4, where the two galaxies actually begin to undergo the merging processes.

% Future Work
Our work shows the importance of considering morphological stage when considering interaction, and that there is a fine interplay between underlying processes and dynamical timescale of an interaction. We require larger samples of correctly staged galaxies to further understand and exploit what these relations are, and how to best investigate when and where star formation and its enhancement occurs. This is also particularly true of the relation between active galactic nuclei and the interaction stage, where we are unable to have the sample size to definitively draw conclusions from our results.
